---
title: "Reproducible Research Course Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and Preprocessing the data

Load data from the local directory into the object "activity", and show the summary of its statistics.

```{r data_reading, cache = TRUE}
activity <- read.csv('./data/activity.csv',stringsAsFactors = FALSE)
summary(activity)
```

Transform the variable, date, into POSIXct format using lubridate package.

```{r date_transformation, message=FALSE}
library(lubridate)
activity$date <- ymd(activity$date)
str(activity)
```

## What is the mean total number of steps?

The total number of steps taken every day is listed as follows,

```{r results='hide'}
require(dplyr)
```

```{r}
avg_steps <- group_by(activity,date) %>% summarize(totalSteps = sum(steps,na.rm=TRUE), na =
                                                       sum(is.na(steps)))
avg_steps
```

We visualize the total number of steps taken each day using a histogram,

```{r histogram}
hist(avg_steps$totalSteps, breaks =20, col='red', main = 'Total number of steps taken per day',xlab = 'Total number of steps taken per day')
```

Furthermore, we are able to derive the mean and median of total steps taken per day.

```{r}
mean_steps <- mean(avg_steps$totalSteps, na.rm = TRUE)
median_steps <- median(avg_steps$totalSteps, na.rm =TRUE)
```

The mean and median of total steps per day are `r round(mean_steps)` and `r median_steps` respectively.

## What is the average daily activity pattern?

In this section, we look at the daily activity pattern using 5-minute interval averaged across all days. First, we group the dataset by 5-minute interval, and present the average steps in a table,

```{r dailyPattern}
daily_pattern <- group_by(activity, interval) %>% summarise(avg_interval = mean(steps,na.rm= TRUE))
daily_pattern

```

As a result, we are able to visualize the table through a time series plot,

```{r}
plot(daily_pattern$interval, daily_pattern$avg_interval, type='l',col='blue', xlab = '5-minute interval', ylab = 'Average 5-minute steps across all days', main = 'Average daily activity pattern')

```

Next, we see which maximum 5-minute interval contains the maximum number of steps averaged across all days, and the corresponding figure.

```{r}
max_interval <- filter(daily_pattern, avg_interval == max(avg_interval))
max_interval
```

In summary, interval "`r max_interval[1]`" contains the maximum number of steps (`r round(max_interval[2])`).

## Inputting missing values

In this section, we will attempt to deal with missing values. First, we look at how many missing values of steps there are in the original dataset.

```{r}
num_na <- summarize(avg_steps,sum(na))
```

There are `r num_na[1,1]` missing values (i.e., total number of rows with NAs) in total in the dataset.

Next, we devise a strategy to fill in these missing values. As there are a large amount of missing values, which concentrate on eight specific days, we choose to utilize the average number of 5-minute steps. In doing so, the resulting dataset would be more informative of the daily activity pattern shown in the last section. We take a bridf look at the modified dataset with an extra variable of steps with missing values filled in.

```{r}
steps_modified <- numeric(nrow(activity))
for (i in 1:nrow(activity)) {
    if (is.na(activity[i,'steps'])) {
        steps_modified[i] <- filter(daily_pattern, interval == activity[i,'interval']) %>% select(avg_interval)
    }
    else {
        steps_modified[i] <- activity[i,'steps']
    }
}

steps_modified <- as.numeric(steps_modified)
activity_modified <- mutate(activity, steps_no_NA = steps_modified)
head(activity_modified)
str(activity_modified)
```

As previously, we draw a histogram of total number of steps taken each day for the modified dataset.

```{r}
avg_steps_modified <- group_by(activity_modified,date) %>% summarize(totalSteps = sum(steps_no_NA,na.rm=TRUE), na = sum(is.na(steps)))
hist(avg_steps_modified$totalSteps,breaks=20,col='magenta', main = 'Total number of steps taken each day', xlab = 'Total number of steps taken each day')
```

Furthermore, we are able to derive the mean and median of total steps taken per day for the modified model.

```{r}
mean_steps_modified <- mean(avg_steps_modified$totalSteps, na.rm = TRUE)
median_steps_modified <- median(avg_steps_modified$totalSteps, na.rm =TRUE)
```

The mean and median of total steps per day for the modified model are `r as.integer(mean_steps_modified)` and `r as.integer(median_steps_modified)` respectively. Compared to the outcome obtained when ignoring missing values, both figures are a bit larger, which makes the analysis more reasonable.

## Are there differences in activity patterns between weekdays and weekends?

In the following, we add a factor variable to the modified dataset in order to determine whether a day is a weekday or weekend.

```{r}
is_weekdays <- rep(0,nrow(activity_modified))
for (i in 1:nrow(activity_modified)) {
    if (weekdays.Date(activity_modified[i,'date'],abbreviate = TRUE) %in% c('Mon','Tue','Wed','Thu','Fri')) {
        is_weekdays[i] <- 'weekday' 
    }
    else {
        is_weekdays[i] <- 'weekend'
    }
}
is_weekdays <- as.factor(is_weekdays)
table(is_weekdays)
activity_modified <- mutate(activity_modified, week = is_weekdays)
str(activity_modified)
```

Last, we observe the daily activity pattern for weekdays and weekends respectively, in time series plots. 

```{r}
daily_pattern_week <- group_by(activity_modified,week,interval) %>% summarize(avg_interval=mean(steps_no_NA))
library(ggplot2)
qplot(interval,avg_interval,data = daily_pattern_week,geom = 'line',facets = week ~ ., main = 'Daily activity pattern for weekdays and weekends', xlab = '5-minute interval')
```